---
title: "ECAAIPAB Mid Term"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list = ls())
```

# LIBRARY

The following libraries are required for this homework.

```{r}
library(tidyverse)
library(ggplot2)
```

# READ DATA

Read the CSV data from `gi_food_data.csv`. Ensure the data file is in the same directory as this markdown R.

```{r}
GI.data <- read.csv("gi_data.csv")
```

Check the table's column names

```{r}
colnames(GI.data)
```

How many columns are there?

```{r}
nrow(GI.data)
```

Answer: 

## Basic Stastics of Data

Use `summary()` function to obtain basic statistics of data such as mean, standard deviation

```{r}
summary(GI.data)
```

Examine the structure of `GI.data`.  We will convert `cho_100g`, `gi_value` and `GL` to float.

```{r}
str(GI.data)
```

Convert columns to numeric type:

```{r}
GI.data$GL <- as.numeric(GI.data$GL)
GI.data$gi_value <- as.numeric(GI.data$gi_value)
GI.data$cho_100g <- as.numeric(GI.data$cho_100g)
```

Convert columns to factor type:

```{r}
GI.data$gl_category <- as.factor(GI.data$gl_category)
GI.data$diogenes_food_grp <- as.factor(GI.data$diogenes_food_grp)
GI.data$source2 <- as.factor(GI.data$source2)
GI.data$diogenes_code <- as.factor(GI.data$diogenes_code)
```

Reexamine the structure:

```{r}
str(GI.data)
```


Which country has the highest `gi_value` (Glycemic Index) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of gi_value
giByCountry <- GI.data %>% filter(!is.na(gi_value)) %>% 
  group_by(country) %>% summarise(mean_gi = mean(gi_value))

# find the country with highest gi_value mean
giByCountry[which.max(giByCountry$mean_gi),]
```
 Answer:
 
Write your code to determine which country has the highest `GL` (Glycemic Load) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of GL

# find the country with highest GL mean

```

Answer: 

Write your code to determine which country has the lowest `GL` (Glycemic Load) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of GL

# find the country with highest GL mean

```

Answer: 

# CORRELATION

First remove NAs from $x$ and $y$:

```{r}
GI.xy <- GI.data %>% select(gi_value, GL) %>% filter(!is.na(gi_value), !is.na(GL))
x <- GI.xy$gi_value
y <- GI.xy$GL
```

Quenstions:

1. What is the the correlation coefficient for `gi_value` and `GL`? 
2. Do they have correlation?
3. What can you say about Glycemic Index and Glycemic Load of food from your results?

```{r}
cor(x, y)
```

Answers:
1.
2.
3.














