---
title: "ECAAIPAB Mid Term"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list = ls())
```

STUDENT NAME:

# PART 1 - BASIC R PROGRAMMING [20 Marks]

The following libraries are required for this homework.

```{r}
library(tidyverse)
library(ggplot2)
```


Read the CSV data from `gi_food_data.csv`. Ensure the data file is in the same directory as this markdown R.

```{r}
GI.data <- read.csv("gi_data.csv", stringsAsFactors=FALSE, fileEncoding="latin1")
```

Check the table's column names

```{r}
colnames(GI.data)
```

*Question 1 [2 Marks]*

How many rows are there in `GI.data`?

```{r}
nrow(GI.data)
```

Answer: 

## Basic Stastics of Data

Use `summary()` function to obtain basic statistics of data such as mean, standard deviation

```{r}
summary(GI.data)
```

Examine the structure of `GI.data`.  We will convert `cho_100g`, `gi_value` and `GL` to float.

```{r}
str(GI.data)
```

Convert columns to numeric type:

```{r}
GI.data$GL <- as.numeric(GI.data$GL)
GI.data$gi_value <- as.numeric(GI.data$gi_value)
GI.data$cho_100g <- as.numeric(GI.data$cho_100g)
```

Convert columns to factor and character types:

```{r}
GI.data$gl_category <- as.factor(GI.data$gl_category)
GI.data$diogenes_food_grp <- as.factor(GI.data$diogenes_food_grp)
GI.data$source2 <- as.character(GI.data$source2)
GI.data$diogenes_code <- as.character(GI.data$diogenes_code)
GI.data$country <- as.factor(GI.data$country)
```

*Question 2 [3 Marks]*
Some of the variables above were converted to numeric, some to factor. What are the benefits of factor type in carrying out statistical analysis? Watch video here for tips 
https://www.youtube.com/watch?v=9f2g7RN5N0I
Answer:

Inspect the types for `GI.data` after conversions.

```{r}
str(GI.data)
```

*Question 3 [3 Marks]* 
What are those countries represented in this dataset?

```{r}
unique(GI.data$country)
```
Answer:


*Question 4 [3 Marks]* 
Which country has the highest `gi_value` (Glycemic Index) mean?  
Remove NAs (NULLs) before applying mean using `filter()`.

```{r}
# group by country and aggregate by mean of gi_value
giByCountry <- GI.data %>% filter(!is.na(gi_value)) %>% 
  group_by(country) %>% summarise(mean_gi = mean(gi_value))

# find the country with highest gi_value mean
giByCountry[which.max(giByCountry$mean_gi),]
```
 Answer:
 
 
*Question 5 [3 Marks]* 
Write your code to determine which country has the highest `GL` (Glycemic Load) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of GL
glByCountry <- GI.data %>% filter(!is.na(GL)) %>% 
  group_by(country) %>% summarise(mean_gl = mean(GL))
# find the country with highest GL mean
giByCountry[which.max(glByCountry$mean_gl),]
```

Answer: 



*Question 6 [3 Marks]* 
Which country has the lowest `gi_value` (Glycemic Index) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of GL
giByCountry <- GI.data %>% filter(!is.na(gi_value)) %>% 
  group_by(country) %>% summarise(mean_gi = mean(gi_value))
# find the country with lowest GI mean
giByCountry[which.min(giByCountry$mean_gi),]
```

Answer: 

*Question 7 [2 Marks]*
Which country has the lowest `GL` (Glycemic Load) mean?  
Remove NAs (NULLs) before applying mean.

```{r}
# group by country and aggregate by mean of GL
glByCountry <- GI.data %>% filter(!is.na(GL)) %>% 
  group_by(country) %>% summarise(mean_gl = mean(GL))

# find the country with highest GL mean
glByCountry[which.min(glByCountry$mean_gl),]
```

Answer: 

# PART 2: STATISTICAL ANALYSIS USING R [30 Marks]

Select `gi_value` and `GL` from `GI.data`.  Remove NAs from $x$ and $y$. Separate the 2 variables and assign them with $x$ and $y$ respectively:

```{r}
GI.xy <- GI.data %>% select(gi_value, GL) %>% filter(!is.na(gi_value), !is.na(GL))
x <- GI.xy$gi_value
y <- GI.xy$GL
```

R provies the `cor()` function to determine correlation coefficient.

*Quenstions 1 [5 Marks]*
1. What is the the correlation coefficient for `gi_value` and `GL`? 
2. Is this a strong or weak correlation? Tips: Refer to https://en.wikipedia.org/wiki/Pearson_correlation_coefficient 
3. What can you say about correlation between Glycemic Index and Glycemic Load of food from your results?

```{r}
cor(x, y)
```
Answers:
1.
2.
3.


Plot the histogram

```{r}
ggplot(GI.data, aes(x=gi_value)) + geom_histogram(binwidth = 1)
```

*Question 2 [5 Marks]*
How many peaks are there exceeding the count of 500?

Answer:



According to https://beyondtype1.org/glycemic-index-diet-diabetes/ in fighting diabetes, we should eat food with low GI (gi_value).

*Question 3 [5 Marks]*
Find the top 5 food name in english translation having the lowest GI?

To answer this question, select the food name in english translation, `gi_value`, sort `gi_value` in ascending order. take the first 5:

```{r}
lowestGI <- GI.data %>% select(english_transl, gi_value) %>% distinct() %>% arrange(gi_value)
head(lowestGI, 5)
```

Answer:
1.
2.
3.
4.
5.


*Question 4 [5 Marks]* 
Find the top 5 food with the highest Glycemic Index:

```{r}
highestGI <- GI.data %>% select(english_transl, gi_value) %>% distinct() %>% arrange(desc(gi_value))
head(highestGI, 5)
```

Answer:
1.
2.
3.
4.
5.


*Question 5 [5 Marks]* 
Find the top 5 food with the highest GL?

```{r}
highestGL <- GI.data %>% filter(!is.na(english_transl), english_transl != '') %>% select(english_transl, GL) %>% distinct() %>% arrange(desc(GL))
head(highestGL, 5)
```

Answer:
1.
2.
3.
4.
5.

*Question 6 [5 Marks]* 
Find the top 5 food with the lowest GL?

```{r}
lowestGL <- GI.data %>% select(english_transl, GL) %>% distinct() %>% arrange(GL)
head(lowestGL, 5)
```

Answer:
1.
2.
3.
4.
5.


According to https://www.diabetes.co.uk/diet/glycemic-load.html, Glycemic load is a measure that takes into account the amount of carbohydrate in a portion of food together with how quickly it raises blood glucose levels. In other words, for every 100g of a particular food, how much carbohydrate (CHO) does it contain? This corresponds to `cho_100g` column which indicates the amount of CHO per 100g.

# PART 3: MANAGING DIABETES USING GLYCEMIC INDEX AND GLYCEMIC LOAD [30 Marks]

Go to youtube to watch this video at https://www.youtube.com/watch?v=Z-cxMdEvsZM&t=126s

In the video, the speaker has define several levels for GI and GL. We will use his video to add 2 columns `GI.data` to categorize the food in the data.

First, write a function for the definition. This function defines high level as $H$, low level as $L$, medium as $<M$ and unknown as $U$

```{r}
GlycemicLevels <- function(inputVal, lowLevel, highLevel)
{
  if(!is.na(inputVal) && inputVal > highLevel) return("H")
  else if(!is.na(inputVal) && inputVal < lowLevel) return("L")
  else if(!is.na(inputVal) && inputVal >= lowLevel && inputVal <= highLevel) return("M")
  else return("U")
}
```

Add `GI_level` column for Glycemic Index Level and `GL_level` for Glycemic Load Level.

```{r}
GI.data$GI_level <- unlist(lapply(GI.data$gi_value, FUN = GlycemicLevels, lowLevel = 55, highLevel = 70))
GI.data$GL_level <- unlist(lapply(GI.data$GL, FUN = GlycemicLevels, lowLevel = 55, highLevel = 70))
```

Check the values of the 2 new columns:

```{r}
head(select(GI.data, english_transl, GI_level, GL_level))
```

We can now use this data to design a diet plan for diabetes patients. In doing so, we will take the approach of selecting food with lowest GI and GL levels. To further fortify our recommendations, we will add one more column which is an addition of `gi_value` and `GL` and use this value to sort the data.

### RECOMMENDED DIET PLAN FOR DIABETIC PATIENTS

First, filter data to have only level $L$. Then, use mutate() function to add the new column. 

```{r}
lowGIGL <- GI.data %>% filter(GI_level == 'L' & GL_level == 'L') %>% 
  select(english_transl, gi_value, GL) %>%
  mutate(sum_gi_gl = gi_value + GL) %>% distinct() %>%
  arrange(sum_gi_gl)
head(lowGIGL, 10)
```

*Question 1 [6 Marks]* 
Inspect the first 10 rows, which of the 2 records are duplicates?
Answers:
1.
2.

To remove duplicates, convert `english_transl` to lower case and reorder the data in ascending order using `sum_gi_gl` column.

```{r}
lowGIGL.clean <- lowGIGL
lowGIGL.clean$english_transl <- as.character(lowGIGL.clean$english_transl)
lowGIGL.clean$english_transl <- tolower(lowGIGL.clean$english_transl)
head(lowGIGL.clean, 10)
```

Notice `english_transl` is now in lower case.

*Question 2 [6 Marks]* 
What is the last item (10th) in the output above?
Answer:

```{r}
lowGIGL.unique <- lowGIGL.clean %>% distinct()
head(lowGIGL.unique, 10)
```

Duplicates are now removed.

*Question 3 [6 Marks]* 
What is the last item (10th) now?
Answer:


### FOOD TO AVOID

We now determine the food which diabetic patients should avoid. We need to identify food with high GI and high GL.

First, convert `english_transl` to lower case:

```{r}
GI.data$english_transl <- tolower(GI.data$english_transl)
```

Like the previous step, create a new column called `sum_gi_gl`, sort the new column in descending order and retrieve the top 10 highest GI and GL.

```{r}
highGIGL <- GI.data %>% filter(GI_level == 'H' & GL_level == 'H') %>%
  mutate(sum_gi_gl = gi_value + GL) %>% arrange(desc(sum_gi_gl)) %>%
  select(english_transl, gi_value, GL, sum_gi_gl) %>% distinct()
head(highGIGL, 10)
```

As we can see, there are still duplicates in the records. They are `kellogg's cornflakes` and `kelloggs cornflakes`. To remove these duplicates, use `str_replace` function to remove the single quote.

```{r}
highGIGL <- highGIGL %>% mutate(english_transl = str_replace(english_transl, "'", "")) %>% distinct()
head(highGIGL, 10)
```

The results look better now. 

## DISCUSSIONS

*Question 4 [6 Marks]*
We now have some knowledge about what to recommend to the diabetic patients and what to avoid. Discuss briefly about the strength and weaknesses about this approach as well as the data and how well they can be improved. Talk also about the feasibility of this approach in helping diabetic patients.

Write your answer below.

Answer:

*Question 5 [6 Marks]*
By using the results in this exam, do some research over the internet to find out if Glycemic results we obtained is a good indicator of the health of a country's population.

Answer:

# PART 4: MORE INSIGHTS [20 Marks]

*Question 1 [10 Marks]*
Based on what you have learned from this course, suggest several other insights that we might be able to find from the GI data.
Answer:


*Question 2 [10 Marks]*
So far, we have only explored data using R programming. In data science, it's crucial to possess a variety of other programming skills such as python, scala, Matlab, Java, SAS, and so forth. Select a few of them and do a research on their future, strength and weaknesses in realizing the aspirations of data science and AI.

Answer:



# SUBMISSION

Upon completion of this exam, convert it to Microsoft Word using Knit -> Knit to Word. Deadline for submission is 17 AUGUST 2019, 7 PM.




