---
title: "Read and Explore Data"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIBRARIES

Load the required libraries

```{r cars}
library(tidyverse)
library(tidyquant)
library(ggplot2)
library(ggalt)
library(quantmod)
library(DBI)
library(RMariaDB)
```

# VECTORS AND SEQUENCES

Create a vector consisting of integers. Check its type using `typeof` function. What can you conclude? Is the vector has the expected type? If not, how do you get the expected `integer` type?

```{r}
vecInt <- c(1,2,3,4,5,6)
vecInt
length(vecInt)
typeof(vecInt)

vecInt <- as.integer(vecInt)
typeof(vecInt)
```

Create a vector consisting of `float` elements:

```{r}
vecFloat <- c(1.1, 2.3, 3.4, 0.1, 8.7)
vecFloat
```

Create vectors with values of mixed types. Under the hood, R will coerce the vector to a specific type.
If an integer and a string 

```{r}
vecMixed1 <- c(1.2, "abc")
class(vecMixed1)

vecMixed2 <- c("abc", TRUE)
class(vecMixed2)
typeof(vecMixed2)

vecMixed1
vecMixed2
```

# CAST TYPES

One type can be cast to another type in R. An integer can be explicitly cast into a string. Use the following functions to cast:
* `as.integer` : cast a value to integer
* `as.numeric` : cast a value to a number

```{r}
num <- 1
class(num)
is.integer(num)
is.numeric(num)

is.integer(1L)
is.numeric(1L)

num1 <- 1.2
is.numeric(num1)
```


## LIST

```{r}
list1 <- list(1.2, 2.3, 0.4, -9.7, -4.1, 6.8)
list1

list1[[1]]
list1[[2]]
# elements 1 and 2
list1[c(1,2)]
# elements 1 to 3
list1[c(1:3)]
# remove first element
list1[c(-1)]
```

Retrieve the list 2 elements:

```{r}
# first find the number of elements
nElems <- length(list1)
# number of elements to be removed
nRemove <- nElems - 2
startElem <- nRemove + 1
list1[c(startElem, nElems)]
```

Mixed types are not coerced in `list`:

```{r}
mixedList <- list(1.2, 3.4, FALSE, "abc", 8, -1L)
mixedList
class(mixedList)
class(mixedList[[1]])
class(mixedList[[2]])
class(mixedList[[3]])
class(mixedList[[4]])
class(mixedList[[5]])
```

Generate a list:

```{r}
numList <- 0:10
numList
class(numList)
# convert to list
class(list(numList))
```


## SEQUENCES

Sequences can be created using the `seq` function. 

```{r}
# create a sequence from 0 to 10 incresed by 1
seq(0, 10, 1)
seq(from=0, to=10, by=1)
seq(from=0, to=10, by=0.5)
seq(from=10, to=100, by=pi)
```


# IF ELSE

If else conditions in R can be implemented using the `ifelse(test, yes, no)` function. 

If BMI of a person is less than 24

```{r}
BMI <- 25
ifelse(BMI > 24, 'overweight', 'normal')
```

If an overweight person exceeds a particular BMI, he / she becomes obese.

```{r}
BMI <- 29
ifelse(BMI > 24,ifelse(BMI < 26, 'overweight', 'obese'), 'normal')
```


# DATES


# FUNCTIONS

# APPLY


# MATRICES

## CREATE MATRICES

A matrix consists of multiple rows of vectors. Number of columns and rows must be specified in creating a matrix.

Create a matrix of 5 columns and 4 rows (4 x 5)
```{r}
mtx1 <- matrix(nrow = 4, ncol = 5)
mtx1
```

Numbers can be initialized when creating a matrix. Create a 4 x 5 matrix with zero initial values:

```{r}
rows <- 4
cols <- 5
zeros <- replicate(rows * cols, 0)
mtx2 <- matrix(zeros, nrow = rows, ncol = cols)
mtx2
```

Initialize the matrix with sequence from 1 to 20. Numbers can be filled 
* from left to right followed by top to bottom (`byrow = TRUE`, defailt behavior)
* from top to bottom followed by left to right (`byrow = FALSE`)

Fill from top to bottom first:

```{r}
seqNum <- seq(from = 1, to = rows * cols, by = 1)
mtx3 <- matrix(seqNum, nrow = rows, ncol = cols, byrow = FALSE)
mtx3
```

Fill from left to right first:

```{R}
mtx4 <- matrix(seqNum, nrow = rows, ncol = cols, byrow = TRUE)
mtx4
```

## SLICING THE MATRIX

Check the matrix's dimension using `dim()` function:

```{r}
dim(mtx3)
```

The first element of `dim` is the row number and the second element is the column number. 

```{r}
dim(mtx3)[1]
dim(mtx3)[2]
```


Retrieve the value of 3rd row and 2nd column:

```{r}
mtx3[3,2]
```

Retrieve values from the 4th row. 

```{r}
mtx3[4,]
```

Retrieve values from 3rd columns:

```{r}
mtx3[,3]
```

Retrieve values from 4th to 5th columns and from 3rd and 4th rows:

```{r}
mtx3[3:4,4:5]
```

Names can be assigned to rows and columns of the matrix.

Assign names to rows:
```{r}
rownames(mtx3) <- c('r1','r2','r3','r4')
mtx3
```

Assign names to columns:

```{r}
colnames(mtx3) <- c('c1','c2','c3','c4','c5')
mtx3
```

```{r}
mtx3['r1','c5']
```

Retrieve values from 1st to 3rd rows and from 3rd to 4th columns:

```{r}
mtx3[1:3, 4:5]
```

A matrix can be converted into a data table (data frame) using `as.data.frame` function:

```{r}
mtx3.df <- as.data.frame(mtx3)
mtx3.df
```

The above table is an index table where `r1`, `r2`, `r3` and `r4` are row names. We want to convert the index to a table column `c0` by using the `rownames_to_column()` function. `rownames_to_column()` is in the `tibble` library in `tidyverse` package.

```{r}
mtx3.df <- rownames_to_column(mtx3.df, 'c0')
mtx3.df
```


A matrix can also be converted into a data frame using the `tibble` library.

First, import the `tibble` library:

```{r}
library(tibble)
```

```{r}
mtx3.tibble <- as_tibble(mtx3)
mtx3.tibble
```

# PLOT HISTOGRAM

## GENERATE DUMMY DATA

```{r}
library(ggplot2)

set.seed(13)
gender <- factor(rep(c("M", "F"), each = 500))
gender
weight.M <- round(c(rnorm(500, mean = 70, sd = 5)))
weight.M[1:20]
weighth.F <- round(c(rnorm(500, mean = 60, sd = 5)))
```

```{r}
gender.df <- data.frame(gender, weight = c(weight.M, weighth.F))
head(gender.df)
```

BASIC HISTOGRAM

```{r}
ggplot(gender.df, aes(x=weight)) + geom_histogram()
```

Change the width of the bin

```{r}
ggplot(gender.df, aes(x=weight)) +   geom_histogram(binwidth=1)
```


Change color

```{r}
gender.plot <- ggplot(gender.df, aes(x=weight)) +  geom_histogram(color="dark blue", fill="light blue")
gender.plot
```


Add mean line and density plot on the histogram


```{r}
gender.plot + geom_vline(aes(xintercept=mean(weight)), color="blue", linetype="dashed", size=1)
```

## HISTOGRAM WITH DENSITY

The histogram is plotted with density instead of count on y-axis
Overlay with transparent density plot. The value of alpha controls the level of transparency

```{r}
ggplot(gender.df, aes(x=weight)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 30) +
 geom_density(alpha=.2, fill="#FF6666") 
```

Change histogram plot line types and colors

```{r}
ggplot(gender.df, aes(x=weight)) +
  geom_histogram(color="black", fill="lightblue", linetype="dashed")

```

## Change histogram plot colors by groups

Calculate the mean of each group

```{r}
group.mean.df <- gender.df %>% group_by(gender) %>% summarise(groupMean = mean(weight))
head(group.mean.df)
```


## Change histogram plot line colors by groups

```{r}
ggplot(gender.df, aes(x=weight, color=gender)) + geom_histogram(fill="white", bins = 30)
```

```{r}
ggplot(gender.df, aes(x=weight, color=gender)) + 
  geom_histogram(fill="white", alpha=0.5, position="identity", bins=30)
```

Interleaved histograms

```{r}
ggplot(gender.df, aes(x=weight, color=gender)) +
  geom_histogram(fill="white", position="dodge")+
  theme(legend.position="top")
```


```{r}
intercept.plot <- ggplot(gender.df, aes(x=weight, color=gender)) +
    geom_histogram(fill="white", position="dodge") +
    geom_vline(data = group.mean.df, aes(xintercept=groupMean, color=gender),
             linetype="dashed") +
    theme(legend.position="top")

intercept.plot
```

Customize colors


```{r}
intercept.plot + scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))
# Use brewer color palettes
intercept.plot + scale_color_brewer(palette="Dark2")
# Use grey scale
intercept.plot + scale_color_grey() + theme_classic() +
  theme(legend.position="top")
```









