---
title: "Module 3 Lab 3"
author: "SIONG, Jong Hang"
date: "03/04/2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIBRARIES

```{r}
library(tidyverse)
library(arules)
library(arulesViz)
library(knitr)
library(gridExtra)
library(lubridate)
```


```{r}
trans <- read.transactions("BreadBasket_DMS.csv", 
                           format="single", cols=c(3,4), sep=",", rm.duplicates=TRUE)


head(trans)
```
Let’s get an idea of what we’re working with.

## Transaction object
```{r}
# Transaction object
trans
```

## Summary
```{r}
# Summary
summary(trans)
```

## Glimpse
```{r}
# Glimpse
glimpse(trans)
```

## Structure
```{r}
# Structure
str(trans)
```

# **Data Dictionary** 

The data set contains 15.010 observations and the following columns,

* `Date`. Categorical variable that tells us the date of the transactions (YYYY-MM-DD format). The column includes dates from 30/10/2016 to 09/04/2017.

* `Time`. Categorical variable that tells us the time of the transactions (HH:MM:SS format).

* `Transaction`. Quantitative variable that allows us to differentiate the transactions. The rows that share the same value in this field belong to the same transaction, that's why the data set has less transactions than observations. 

* `Item`. Categorical variable containing the products. 

# **Data Analysis**

Before applying the Apriori algorithm on the data set, we are going to show some visualizations to learn more about the transactions. For example, we can generate an `itemFrequencyPlot()` to create an item Frequency Bar Plot to view the distribution of products. 

```{r fig.align='center'}
# Absolute Item Frequency Plot
itemFrequencyPlot(trans, topN=15, type="absolute", col="wheat2",xlab="Item name", 
                  ylab="Frequency (absolute)", main="Absolute Item Frequency Plot")
```
 
The `itemFrequencyPlot()` allows us to show the absolute or relative values. If absolute it will plot numeric frequencies of each item independently. If relative it will plot how many times these items have appeared as compared to others, as it's shown in the following plot. 

```{r fig.align='center'}
# Relative Item Frequency Plot
itemFrequencyPlot(trans, topN=15, type="relative", col="lightcyan2", xlab="Item name", 
                  ylab="Frequency (relative)", main="Relative Item Frequency Plot")
```
   
Coffee is the best-selling product by far, followed by bread and tea. Let's display some other visualizations describing the time distribution using the `ggplot()` function. 

* Transactions per month

* Transactions per weekday

* Transactions per hour 

```{r fig.align='center'}
# Load data 
trans_csv <- read.csv("BreadBasket_DMS.csv")

# Transactions per month
trans_csv %>%
  mutate(Month=as.factor(month(Date))) %>%
  group_by(Month) %>%
  summarise(Transactions=n_distinct(Transaction)) %>%
  ggplot(aes(x=Month, y=Transactions)) +
  geom_bar(stat="identity", fill="mistyrose2", show.legend=FALSE, colour="black") +
  geom_label(aes(label=Transactions)) +
  labs(title="Transactions per month") +
  theme_bw() 
```

The data set includes dates from 30/10/2016 to 09/04/2017, that's why we have so few transactions in October and April. 

```{r fig.align='center'}
# Transactions per weekday
trans_csv %>%
  mutate(WeekDay=as.factor(weekdays(as.Date(Date)))) %>%
  group_by(WeekDay) %>%
  summarise(Transactions=n_distinct(Transaction)) %>%
  ggplot(aes(x=WeekDay, y=Transactions)) +
  geom_bar(stat="identity", fill="peachpuff2", show.legend=FALSE, colour="black") +
  geom_label(aes(label=Transactions)) +
  labs(title="Transactions per weekday") +
  scale_x_discrete(limits=c("Monday", "Tuesday", "Wednesday", "Thursday",
                            "Friday", "Saturday", "Sunday")) +
  theme_bw() 
```

As we can see, Saturday is the busiest day in the bakery. Conversely, Wednesday is the day with fewer transactions.

```{r fig.align='center'}
# Transactions per hour
trans_csv %>%
  mutate(Hour=as.factor(hour(hms(Time)))) %>%
  group_by(Hour) %>%
  summarise(Transactions=n_distinct(Transaction)) %>%
  ggplot(aes(x=Hour, y=Transactions)) +
  geom_bar(stat="identity", fill="steelblue1", show.legend=FALSE, colour="black") +
  geom_label(aes(label=Transactions)) +
  labs(title="Transactions per hour") +
  theme_bw()
```

There’s not much to discuss with this visualization. The results are logical and expected.

# **Apriori algorithm**

## Choice of support and confidence

The first step in order to create a set of association rules is to determine the optimal thresholds for support and confidence. If we set these values too low, then the algorithm will take longer to execute and we will get a lot of rules (most of them will not be useful). Then, what values do we choose?
We can try different values of support and confidence and see graphically how many rules are generated for each combination.

```{r results='hide'}
# Support and confidence values
supportLevels <- c(0.1, 0.05, 0.01, 0.005)
confidenceLevels <- c(0.9, 0.8, 0.7, 0.6, 0.5, 0.4, 0.3, 0.2, 0.1)

# Empty integers 
rules_sup10 <- integer(length=9)
rules_sup5 <- integer(length=9)
rules_sup1 <- integer(length=9)
rules_sup0.5 <- integer(length=9)

# Apriori algorithm with a support level of 10%
for (i in 1:length(confidenceLevels)) {
  
  rules_sup10[i] <- length(apriori(trans, parameter=list(sup=supportLevels[1], 
                                   conf=confidenceLevels[i], target="rules")))
  
}

# Apriori algorithm with a support level of 5%
for (i in 1:length(confidenceLevels)) {
  
  rules_sup5[i] <- length(apriori(trans, parameter=list(sup=supportLevels[2], 
                                  conf=confidenceLevels[i], target="rules")))
  
}

# Apriori algorithm with a support level of 1%
for (i in 1:length(confidenceLevels)) {
  
  rules_sup1[i] <- length(apriori(trans, parameter=list(sup=supportLevels[3], 
                                  conf=confidenceLevels[i], target="rules")))
  
}

# Apriori algorithm with a support level of 0.5%
for (i in 1:length(confidenceLevels)) {
  
  rules_sup0.5[i] <- length(apriori(trans, parameter=list(sup=supportLevels[4], 
                                    conf=confidenceLevels[i], target="rules")))
  
}
```


In the following graphs we can see the number of rules generated with a support level of 10%, 5%, 1% and 0.5%. 

```{r fig.align='center'}
# Number of rules found with a support level of 10%
plot1 <- qplot(confidenceLevels, rules_sup10, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 10%") +
  theme_bw()

# Number of rules found with a support level of 5%
plot2 <- qplot(confidenceLevels, rules_sup5, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 5%") + 
  scale_y_continuous(breaks=seq(0, 10, 2)) +
  theme_bw()

# Number of rules found with a support level of 1%
plot3 <- qplot(confidenceLevels, rules_sup1, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 1%") + 
  scale_y_continuous(breaks=seq(0, 50, 10)) +
  theme_bw()

# Number of rules found with a support level of 0.5%
plot4 <- qplot(confidenceLevels, rules_sup0.5, geom=c("point", "line"), 
               xlab="Confidence level", ylab="Number of rules found", 
               main="Apriori with a support level of 0.5%") + 
  scale_y_continuous(breaks=seq(0, 130, 20)) +
  theme_bw()

# Subplot
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```
  
We can join the four lines to improve the visualization.

```{r fig.align='center'}
# Data frame
num_rules <- data.frame(rules_sup10, rules_sup5, rules_sup1, rules_sup0.5, confidenceLevels)

# Number of rules found with a support level of 10%, 5%, 1% and 0.5%
ggplot(data=num_rules, aes(x=confidenceLevels)) +
  
  # Plot line and points (support level of 10%)
  geom_line(aes(y=rules_sup10, colour="Support level of 10%")) + 
  geom_point(aes(y=rules_sup10, colour="Support level of 10%")) +
  
  # Plot line and points (support level of 5%)
  geom_line(aes(y=rules_sup5, colour="Support level of 5%")) +
  geom_point(aes(y=rules_sup5, colour="Support level of 5%")) +
  
  # Plot line and points (support level of 1%)
  geom_line(aes(y=rules_sup1, colour="Support level of 1%")) + 
  geom_point(aes(y=rules_sup1, colour="Support level of 1%")) +
  
  # Plot line and points (support level of 0.5%)
  geom_line(aes(y=rules_sup0.5, colour="Support level of 0.5%")) +
  geom_point(aes(y=rules_sup0.5, colour="Support level of 0.5%")) +
  
  # Labs and theme
  labs(x="Confidence levels", y="Number of rules found", 
       title="Apriori algorithm with different support levels") +
  theme_bw() +
  theme(legend.title=element_blank())
```


Let's analyze the results,

* **Support level of 10%**. We only identify a few rules with very low confidence levels. This means that there are no relatively frequent associations in our data set. We can't choose this value, the resulting rules are unrepresentative.

* **Support level of 5%**. We only identify a rule with a confidence of at least 50%. It seems that we have to look for support levels below 5% to obtain a greater number of rules with a reasonable confidence. 

* **Support level of 1%**. We started to get dozens of rules, of which 13 have a confidence of at least 50%.

* **Support level of 0.5%**. Too many rules to analyze!

To sum up, we are going to use a support level of 1% and a confidence level of 50%. 

## Execution

Let's execute the Apriori algorithm with the values obtained in the previous section.

```{r results='hide'}
# Apriori algorithm execution with a support level of 1% and a confidence level of 50%
rules_sup1_conf50 <- apriori(trans, parameter=list(sup=supportLevels[3], 
                             conf=confidenceLevels[5], target="rules"))
```

The generated association rules are the following,

```{r}
# Association rules
inspect(rules_sup1_conf50)
```

We can also create an HTML table widget using the `inspectDT()` function from the `aruslesViz` package. Rules can be interactively filtered and sorted.

How do we interpret these rules? 

* 52% of the customers who bought a hot chocolate algo bought a coffee.

* 63% of the customers who bought a spanish brunch also bought a coffee.

* 73% of the customers who bought a toast also bought a coffee.

And so on. It seems that in this bakery there are many coffee lovers.

## Visualize association rules

We are going to use the `arulesViz` package to create the visualizations. Let's begin with a simple scatter plot with different measures of interestingness on the axes (lift and support) and a third measure (confidence) represented by the color of the points. 

```{r fig.align='center'}
# Scatter plot
plot(rules_sup1_conf50, measure=c("support","lift"), shading="confidence")
```
 
The following visualization represents the rules as a graph with items as labeled vertices, and rules represented as vertices connected to items using arrows.

```{r fig.align='center'}
# Graph
plot(rules_sup1_conf50, method="graph")
```
 
We can also change the graph layout.  

```{r fig.align='center'}
# Graph
plot(rules_sup1_conf50, method="graph", control=list(layout=igraph::in_circle()))
```

What else can we do? We can represent the rules as a grouped matrix-based visualization. The support and lift measures are represented by the size and color of the ballons, respectively. 
In this case it's not a very useful visualization, since we only have coffe on the right-hand-side of the rules. 

```{r fig.align='center'}
# Grouped matrix plot
plot(rules_sup1_conf50, method="grouped")
```
 
There's an awesome function called `ruleExplorer()` that explores association rules using interactive manipulations and visualization using shiny. Unfortunately, R Markdown still doesn't support shiny app objects. 


## Another execution {.tabset .tabset-fade .tabset-pills}

We have executed the Apriori algorithm with the appropriate support and confidence values. What happens if we execute it with low values? How do the visualizations change? Let's try with a support level of 0.5% and a confidence level of 10%.

```{r results='hide'}
# Apriori algorithm execution with a support level of 0.5% and a confidence level of 10%
rules_sup0.5_conf10 <- apriori(trans, parameter=list(sup=supportLevels[4], conf=confidenceLevels[9], target="rules"))
```

It's impossible to analyze these visualizations! For larger rule sets visual analysis becomes difficult. Furthermore, most of the rules are useless. That's why we have to carefully select the right values of support and confidence. 

### Graph

```{r fig.align='center', message=FALSE, warning=FALSE}
# Graph
plot(rules_sup0.5_conf10, method="graph", control=list(layout=igraph::in_circle()))
```
  
### Parallel coordinates plot

```{r fig.align='center'}
# Parallel coordinates plot
plot(rules_sup0.5_conf10, method="paracoord", control=list(reorder=TRUE))
```
  
### Grouped matrix plot

```{r fig.align='center'}
# Grouped matrix plot
plot(rules_sup0.5_conf10, method="grouped")
```

### Scatter plot

```{r fig.align='center'}
# Scatter plot
plot(rules_sup0.5_conf10, measure=c("support","lift"), shading="confidence", jitter=0)
```





