---
title: "Stock Forecast ARIMA"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LIBRARY

```{r}
library(tidyverse)
library(ggplot2)
library(forecast)
```

## Including Plots

You can also embed plots, for example:

```{r}
stock.data <- read.table("IBMdaily.txt",header = T)
stock.data$Date <- as.Date(as.character(stock.data$Date), format="%m/%d/%Y")
attach(stock.data)
```

Inspect Data

```{r}
head(stock.data)
```

# Events

```{r}
evt.ficom <- which(Date=="1964-04-07")
evt.loss <- which(Date=="1993-07-27")
evt.lenovo <- which(Date=="2005-05-02")
evt.healthm <- which(Date=="2015-08-05")
evt.truven <- which(Date=="2016-02-18")
evt.alldates <- c(evt.ficom, evt.loss, evt.lenovo, evt.healthm, evt.truven)
```

# VISUALIZATION

ˆ
```{r}
ggplot(stock.data, aes(x = Date, y = AdjClose)) + 
  geom_line(color = "red") + xlab("Date") + 
  ylab("Stock Close Price") + 
  ggtitle("Stock Close Price") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_vline(xintercept = as.numeric(Date[evt.alldates]), color = c("blue"), 
             linetype = "dashed")
```

## IBM Stock Price: Non-constant variance => Transform

```{r}
AdjClose.tr = log(AdjClose)
ggplot(stock.data, aes(Date, AdjClose.tr)) + geom_line() + xlab("") + ylab("Stock Close Price")+
geom_vline(xintercept = as.numeric(Date[evt.alldates]))
```

```{r}
ts.price = ts(rev(AdjClose.tr),start=c(1962,1,2),frequency=365.25)
```

# Differencing to Remove Trend

```{r}
diff.ts.price = diff(ts.price)
acf(as.vector(diff.ts.price), main="ACF: One-Lag Difference Transf Price")
pacf(diff.ts.price, main="PACF: One-Lag Difference Transf Price")
```

```{r}
ggPacf(diff.ts.price, lag.max = 40, plot = TRUE, na.action = na.contiguous)
```


## Apply ARIMA

## Order selection -- AIC 

```{r}
n = length(ts.price)
norder = 6
p = c(1:norder)-1; q = c(1:norder)-1
aic = matrix(0,norder,norder)
for(i in 1:norder){
   for(j in 1:norder){
    modij = arima(ts.price,order = c(p[i],1,q[j]), method='ML')
    aic[i,j] = modij$aic-2*(p[i]+q[j]+1)+2*(p[i]+q[j]+1)*n/(n-p[i]-q[j]-2)
   }  
}
```


```{r}
aicv = as.vector(aic)  
plot(aicv,ylab="AIC values")
indexp = rep(c(1:norder),norder)
indexq = rep(c(1:norder),each=norder)
indexaic = which(aicv == min(aicv))
porder = indexp[indexaic]-1
qorder = indexq[indexaic]-1
```
```{r}
final_model = arima(ts.price, order = c(porder,1,qorder), method = "ML")
```

## GOF: residual analysis

```{r}
par(mfrow=c(2,2))
plot(resid(final_model), ylab='Residuals',type='o',main="Residual Plot")
abline(h=0)
acf(resid(final_model),main="ACF: Residuals")
hist(resid(final_model),xlab='Residuals',main='Histogram: Residuals')
qqnorm(resid(final_model),ylab="Sample Q",xlab="Theoretical Q")
qqline(resid(final_model))

Box.test(final_model$resid, lag = (porder+qorder+1), type = "Box-Pierce", fitdf = (porder+qorder))
Box.test(final_model$resid, lag = (porder+qorder+1), type = "Ljung-Box", fitdf = (porder+qorder))
```


# FORECASTING WITH ARIMA

## 10 Days Ahead: 

```{r}
n <- length(ts.price)
nfit <- n - 10
outprice <- arima(ts.price[1:nfit], order = c(porder,1,qorder),method = "ML")
outpred <- predict(outprice,n.ahead=10)
ubound <- outpred$pred + 1.96*outpred$se
lbound <- outpred$pred - 1.96*outpred$se
ymin <- min(exp(lbound))
ymax <- max(exp(ubound))
```

```{r}
plot(rev(stock.data$Date)[(n-50):n],exp(ts.price[(n-50):n]),type="l", ylim=c(ymin,ymax), xlab="Time", ylab="Stock Price")
points(rev(stock.data$Date)[(nfit+1):n],exp(outpred$pred),col="red")
lines(rev(stock.data$Date)[(nfit+1):n],exp(ubound),lty=3,lwd= 2, col="blue")
lines(rev(stock.data$Date)[(nfit+1):n],exp(lbound),lty=3,lwd= 2, col="blue")
```


# Compute Accuracy Measures

```{r}
obsprice = exp(ts.price[(nfit+1):n]) 
predprice = exp(outpred$pred)
```

## Mean Squared Prediction Error (MSPE)

```{r}
mean((predprice-obsprice)^2)
```


## Mean Absolute Prediction Error (MAE)

```{r}
mean(abs(predprice-obsprice))
```

## Mean Absolute Percentage Error (MAPE)

```{r}
mean(abs(predprice-obsprice)/obsprice)
```

## Precision Measure (PM)

```{r}
sum((predprice-obsprice)^2)/sum((obsprice-mean(obsprice))^2)
```

### Does the observed data fall outside the prediction intervals?

```{r}
sum(obsprice<exp(lbound))+sum(obsprice>exp(ubound))
```

## Daily Prediction over a period of 10 days

```{r}
outpred.10 = NULL
ubound.10 = NULL
lbound.10 = NULL
n = length(ts.price)
for(i in 1:10){
   nfit = n-(10-i-1)
   outprice = arima(ts.price[1:nfit], order = c(porder,1,qorder),method = "ML")
   pred.1 = predict(outprice,n.ahead=1)
   outpred.10 = c(outpred.10,pred.1$pred)
   ubound.10 = c(ubound.10, pred.1$pred+1.96*pred.1$se)
   lbound.10 = c(lbound.10, pred.1$pred-1.96*pred.1$se)
  }
```

## Compute Accuracy Measures

```{r}
predprice.10 = exp(outpred.10)
```


### Mean Squared Prediction Error (MSPE)
```{r}
mean((predprice.10-obsprice)^2)
```

### Mean Absolute Prediction Error (MAE)

```{r}
mean(abs(predprice.10-obsprice))
```

### Mean Absolute Percentage Error (MAPE)

```{r}
mean(abs(predprice.10-obsprice)/obsprice)
```


### Precision Measure (PM)

```{r}
sum((predprice.10-obsprice)^2)/sum((obsprice-mean(obsprice))^2)
```

### Does the observed data fall outside the prediction intervals?

```{r}
sum(obsprice<exp(lbound.10))+sum(obsprice>exp(ubound.10))
```

```{r}
nfit = n-10
ymin = min(exp(c(lbound,lbound.10)))
ymax = max(exp(c(ubound,ubound.10)))
plot(rev(stock.data$Date)[(n-30):n],exp(ts.price[(n-30):n]),type="l", ylim=c(ymin,ymax), xlab="Time", ylab="Stock Price")
points(rev(stock.data$Date)[(nfit+1):n],exp(outpred$pred),col="red")
lines(rev(stock.data$Date)[(nfit+1):n],exp(ubound),lty=3,lwd= 2, col="blue")
lines(rev(stock.data$Date)[(nfit+1):n],exp(lbound),lty=3,lwd= 2, col="blue")
points(rev(stock.data$Date)[(nfit+1):n],exp(outpred.10),col="green")
lines(rev(stock.data$Date)[(nfit+1):n],exp(ubound.10),lty=3,lwd= 2, col="purple")
lines(rev(stock.data$Date)[(nfit+1):n],exp(lbound.10),lty=3,lwd= 2, col="purple")
legend(rev(stock.data$Date)[n-25],170,legend=c("1 Lag Ahead Prediction", "1 Lag Ahead Prediction Interval",
"10 Lags Ahead Prediction","10 Lags Ahead Prediction Interval" ),col=c("green","purple","red","blue"),pch=1,lty = c(0,3,0,3))

```

