---
title: "Read and Explore Data"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIBRARIES

Load the required libraries

```{r cars}
library(tidyverse)
library(tidyquant)
library(ggplot2)
library(ggalt)
library(quantmod)
library(DBI)
library(RMariaDB)
```

## HELPER FUNCTIONS

### FUNCTION TO READ CSV FILES

The Data Path has the directory of pattern `/<home>/data/<data dir>/<data file>`

```{r}
DataFile <- function(dataDir, dataFilename)
{
    home.dir <- Sys.getenv('HOME')
    data.path <- paste(home.dir,'data', dataDir, dataFilename, sep = '/')
    data <- read.csv(data.path)
    data
}

```

## READ DATA FILES

```{r}
airlinesDirName <- 'airlines_data'
flights.data <- DataFile(airlinesDirName,'flights.csv')
airlines.data <- DataFile(airlinesDirName, 'airlines.csv')
airports.data <- DataFile(airlinesDirName, 'airports.csv')

```

Inspect Flights data

```{r}
head(flights.data)
```

Inspect Airlines Data

```{r}
head(airlines.data)
```

Inspect Airports Data

```{r}
head(airports.data)
```


# READ MYSQL DATA

```{r}
learn.db <- dbConnect(RMariaDB::MariaDB(),
                      user = "learninguser",
                      password = "learning",
                      dbname = "learningdb",
                      host = "localhost")
```

```{r}
dbGetQuery(learn.db, "SHOW TABLES")
```

```{r}
food.df <- dbGetQuery(learn.db, "SELECT * FROM gi_food_data")
head(food.df)
```

# BUILT-IN DATA

R installation comes with a number of datasets. Use `data()` to display the list of datasets.

```{r}
data()
```

## MTCARS DATASET

```{r}
data(mtcars)
mtcars <- as.data.frame(mtcars)
head(mtcars)
```


## DISPLAY SETTING

* Turn off scientific exponential (scipen = 999)
* Set plot's width and height to 4 and 3

```{r}
options(scipen = 999, repr.plot.width=4, repr.plot.height=3) # turn-off scientific notation like 1e+48
library(ggplot2)
theme_set(theme_bw()) # pre-set the bw theme.
```

### MTCARS DATA EXPLORATION

Plot Points

```{r}
scatter.plot <- ggplot(mtcars, aes(x=mpg, y=hp)) + geom_point()
scatter.plot
```

Resize the Points by `Displacement (disp)`

```{r}
scatter.plot <- scatter.plot + geom_point(aes(col=cyl, size=disp))
scatter.plot
```

Add Smooth Line using LOESS algorithm

```{r}
scatter.plot <- scatter.plot + geom_smooth(method="loess", se=F)
scatter.plot
```


Extend the X and Y Axis Limits a little

```{r}
scatter.plot <- scatter.plot + xlim(c(10, 35)) + ylim(c(0, 350))
scatter.plot
```


Complete with Title and Labels

```{r}
scatter.plot <- scatter.plot + labs(subtitle="HP Vs MPG", y="HP", x="MPG", 
                                    title="Scatterplot", caption = "Source:R")
scatter.plot
```

```{r}
scatter.plot <- ggplot(mtcars, aes(x = mpg, y = hp)) + 
    geom_point(aes(col = cyl, size=disp)) +
    geom_smooth(method="loess", se = FALSE) +
    labs(subtitle="HP Vs MPG", y = "HP", x = "MPG",
         title="Scatterplot", caption = "Source: R")
scatter.plot
```

### SCATTER PLOT WITH ENCIRCLING

```{r}
mtcars.circle <- mtcars[mtcars$hp >= 230,]
```

```{r}
mtcars.plot <-  ggplot(mtcars, aes(x = mpg, y = hp)) + 
                geom_point(aes(col = cyl, size = disp)) + # draw points
                geom_smooth(method="loess", se = FALSE)
mtcars.plot
```

Add Circle

```{r}
mtcars.plot <- mtcars.plot + 
               geom_encircle(aes(x = mpg, y = hp),
                             data = mtcars.circle, color = "red", 
                             size = 2, expand = 0.08) +
               labs(subtitle="Area Vs Population", y="Population", x="Area", 
                    title="Scatterplot + Encircle", caption="Source: midwest")
mtcars.plot
```

# STREAM FINANCIAL DATA

Stream the following stocks of Aircraft Manufacturing sector:

* Boeing (symbol: BA)
* Airbus (symbol: AIR)
* Lockheed Martin (symbol: LMT)

```{r}
BA <- tq_get(x = "BA")
AIR <- tq_get(x = "AIR")
LMT <- tq_get(x = "LMT")
```

The first few lines of BA data:

```{r}
head(BA)
```

The earliest and latest dates:

```{r}
min(BA$date)
max(BA$date)
nrow(BA)
```

Get data from 2018 until today. First, set today's date object

```{r}
today <- now()
today

# take only the date portion
today <- as.Date(today)
today
```


```{r}
BA.from2018 <- tq_get(x = "BA", from= '2018-01-01', to = today)
AIR.from2018 <- tq_get(x = "AIR", from= '2018-01-01', to = today)
LMT.from2018 <- tq_get(x = "LMT", from= '2018-01-01', to = today)
```

```{r}
nrow(BA.from2018)
min(BA.from2018$date)
max(BA.from2018$date)
```

## OTHER COUNTRY'S STOCK

Stream Singapore's Keppel Stock (symbol: BN4). 

```{r}
BN4.from2018 <- tq_get(x = "BN4.SI", from= '2018-01-01', to = today)
head(BN4.from2018)
tail(BN4.from2018)
```

Stream Hng Kong's Cheung Kong's Stock (symbol: 0001.HK)

```{r}
CKH.from2018 <- tq_get(x = "0001.HK",  from = '2018-01-01', to = today)
head(CKH.from2018)
tail(CKH.from2018)
```


```{r}
MAYBANK.from2018 <- tq_get(x = "1155.KL",  from = '2018-01-01', to = today)
head(MAYBANK.from2018)
```


# DATA OPTIONS

There are other data apart from stock prices.

```{r}
tq_get_options()
```

# STOCK EXCHANGES

Stock exchanges available in tidyquant

```{r}
tq_exchange_options()
```


Connect to NYSE

```{r}
NYSE <- tq_exchange("nyse")
```

Examine the structure of NYSE

```{r}
str(NYSE)
```

```{r}
head(NYSE)
tail(NYSE)
```

# DATA PROCESSING USING dplyr 

The library we use to transform data is `dplyr` in `tidyverse` package.

We will perform some data processing on the *airlines* data.

```{r}
head(flights.data)
```

Create a function to combine `YEAR`, `MONTH` and `DAY` columns into a single field of `DATE`. 

```{r}
buildDateFunc <- function(year, month, day)
{
    dateStr <- paste(as.character(year),as.character(month),as.character(day), sep = "/")
    dateObj <- as.Date(dateStr, "%Y/%m/%d")
    dateObj
}
```

Apply the `buildDateFunc` to flighs data. This can be done by mutating the data frame using `mutate` function of `dplyr`. Name the new date field as `FLIGHT_DATE`.

```{r}
flights.data.date <- flights.data %>% mutate(FLIGHT_DATE = buildDateFunc(YEAR, MONTH, DAY))
head(flights.data.date)
str(flights.data.date)
```

Retrieve the basic statistics of flights data.


```{r}
summary(flights.data.date)
```

Get the earliest and latest `FLIGHT_DATE`.

```{r}
min(flights.data.date$FLIGHT_DATE)
max(flights.data.date$FLIGHT_DATE)
```

## Filter Data

Retrieve flights of month 2 and day 2

```{r}
flights_m2_d2 <- flights.data.date %>% filter(MONTH == 2, DAY == 2)
head(flights_m2_d2)
nrow(flights_m2_d2)
```

Alternatively:

```{r}
flights_m2_d2_alt <- flights.data.date %>% 
    filter(month(FLIGHT_DATE) == 2, day(FLIGHT_DATE) == 2)
head(flights_m2_d2_alt)
nrow(flights_m2_d2_alt)
```

Use logical AND

```{r}
flights_m2_d2_log <- flights.data.date %>% 
    filter(month(FLIGHT_DATE) == 2 & day(FLIGHT_DATE) == 2)
head(flights_m2_d2_log)
nrow(flights_m2_d2_log)
```

Use Logical OR to filter flights of `MONTH` 2 and 6

```{r}
flights_m2_m6_log <- flights.data.date %>% 
    filter(month(FLIGHT_DATE) == 2 | month(FLIGHT_DATE) == 6)
head(flights_m2_m6_log)
nrow(flights_m2_m6_log)
```

Verify only months 2 and 6 are retrieved.

```{r}
unique(flights_m2_m6_log$MONTH)
```

## SORT DATA

Function `arrange` in `dplyr` is used to sort data. By default, the sorting occurs in ascending order. Use `arrange(desc(fieldname))` to sort in descending order.

For now, we are only interested in `TAXI_IN`, `FLIGHT_DATE`, `DESTINATION_AIRPORT`, `AIRLINE`. We want to find out on which days the arrival airlines took the longest time to taxi in and at which airports.

First, select only the desirable columns:

```{r}
flights.taxi_in <- select(flights.data.date, FLIGHT_DATE, 
                          AIRLINE, DESTINATION_AIRPORT,
                          TAXI_IN)
flights.taxi_in.sorted <- flights.taxi_in %>% arrange(desc(TAXI_IN))

head(flights.taxi_in.sorted)

str(flights.taxi_in)
```

The Taxi in time is in seconds. Add a column to see the time in minutes:

```{r}
flights.taxi_in.sorted.minutes <- flights.taxi_in.sorted %>% mutate(TAXI_IN_MIN = round(TAXI_IN/60, 2))
head(flights.taxi_in.sorted.minutes)
```

# DATA WRANGLING

## Glycemic Index Data of Food

This exercise illustrates data wrangling to combine several CSV files into a single dataset. The source of the data files is http://www.diogenes-eu.org/GI-Database/Default.htm

The CSV files are:

* Diogenes-GI-Database-for-Website-Czech.csv
* Diogenes-GI-Database-for-Website-Danish.csv
* Diogenes-GI-Database-for-Website-Dutch.csv
* Diogenes-GI-Database-for-Website-German.csv
* Diogenes-GI-Database-for-Website-Greek.csv
* Diogenes-GI-Database-for-Website-Spanish.csv
* Diogenes-GI-Database-for-Website-UK.csv

```{r}
rm(list = ls())

HOME <- Sys.getenv('HOME')
GI.Data.Dir <- paste(HOME, 'Data', 'food', sep = '/')
GI.Data.Dir
```

Read Data Files

```{r}
GI.csv.files <- list.files(GI.Data.Dir, pattern = 'Diogenes-.*\\.csv', full.names = TRUE)
GI.csv.files

GI.cols <- c('diogenes_code','country_food_code',
             'food_name','english_transl',
             'diogenes_food_grp','cho_100g',
             'gi_value','GL', 'gl_category',
             'source1','source2')

GI.czech <- read.csv(GI.csv.files[1])
colnames(GI.czech) <- GI.cols
head(GI.czech)
```


```{r}

build_gi_food_data <- function(fullPath)
{
    GI.cols <- c('diogenes_code','country_food_code',
             'food_name','english_transl',
             'diogenes_food_grp','cho_100g',
             'gi_value','GL', 'gl_category',
             'source1','source2','comment')
    data <- read.csv(fullPath, )[,1:12]
    colnames(data) <- GI.cols
    data <- data %>% mutate_at(vars(starts_with("source")),funs(factor))
    data <- data %>% mutate_at(vars(starts_with("cho_")),funs(factor))
    data <- data %>% mutate_at(vars(starts_with("GL")),funs(factor))
    data
}

head(build_gi_food_data(GI.csv.files[5]))

GI.data.all <- rbind_list(lapply(GI.csv.files, build_gi_food_data))
head(GI.data.all)
```








